A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-02-19 analysis/paper/paper.Rmd
update paper

```{r setup, include=FALSE}
source("setup.R")
```
`r glue::collapse(sample(c("Wouter Saelens* ¹ ²", "Robrecht Cannoodt* ¹ ² ³")), ", ")`, Helena Todorov¹ ² ⁴ ⁵, Yvan Saeys¹ ²

@@

Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes. Since 2014, more than 50 trajectory inference methods have been developed, each with its own set of methodological characteristics. This makes a direct comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking. We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard. Using this framework, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets. We employ several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness. We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods. Our evaluation pipeline is fully reproducible, can be easily extended, and is available at https://www.github.com/dynverse.

@@

Single-cell -omics technologies now make it possible to more accurately model biological systems than ever before. One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation. When cells are sampled from a population in which cells are at different unknown points in the dynamic process, trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016]. Because they offer an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification of new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs. This both for specific subsets of cells, but in the near future also for whole tissues or organisms [@regev_scienceforumhuman_2017].

@@

Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`). Initially, most trajectory inference methods fixed the trajectory topology to a linear `r cite_methods(fixed_linear)`, bifurcating `r cite_methods(fixed_bifurcating)` or other `r cite_methods(fixed_other)` topology, by either algorithmic limitations `r cite_methods(fixed_algorithm)` or through user parameters `r cite_methods(fixed_parameters)`. The focus of these methods was therefore mainly to correctly ordering the cells along this fixed topology. However, several recent methods now infer both the ordering and the topology at the same time, although some topological limitations are still imposed (`r ref("sfig", "methods_over_time", "b")`). Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].

<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
Although trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others. Furthermore, with the ongoing strong scalability in single-cell -omics technologies, it becomes more and more important that analysis methods become both scalable [@aibar_scenicsinglecellregulatory_2017] and user friendly.

<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem. Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art. In this study, we therefore for the first time developed a comprehensive evaluation framework for trajectory inference methods. We test both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge. We include a rigorous parameter optimisation, making sure that parameters are optimized while avoiding overfitting on characteristics of specific datasets. To make our framework reusable and extendable, we make use of continuous analysis [@beaulieu-jones_reproducibility_2017], which allows the developers of new methods to easily test their methods and compare them against the state-of-the-art.

@@

We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface (`r ref("table", "methods")`). We characterized all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characterisation")`).

@@

At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characterisation", "a")`). However, only a handful, mostly recent, methods can currently handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).

@@

`r anchor("fig", "method_characterisation")`
![](`r figure_file("method_characteristics.svg", experiment_id="method_characterisation")`)
`r ref("fig", "method_characterisation", anchor=TRUE)` First characterization of each method.

@@

Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characterisation", "b")`). Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`). However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characterisation", "e")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characterisation", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].

@@

The underlying algorithms can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building steps. Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "method_components")`). MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "method_components", "b")`).

@@

We wrapped each method and postprocessed its output into a common probabilistic format. This format works at two level: at the To evaluate the accuracy of each method, we gathered both synthetic and real datasets, and used these to assess how good a method is at correctly ordering cells, detecting the right trajectory topology and 2
# Discussion
<!-- Some intro -->
Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data. Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] created by Sean Davis and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) . It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice. In this preprint, we summarise the initial version of our evaluation of these methods, focussing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.

@@

* DPT: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point. This was then processed as described above.
* Sincell: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.  This was then processed as described above.
* SLICER: Although this method returns a branch network and a global pseudotime, it was not clear to us how to assign cells to individual branches in the network. We therefore used the cell graph provided by the method to calculate the shortest paths between each branch, and then projected other cells onto this backbone graph structure, creating a local pseudotime for every cell within its branch. The output was then further processed as above.
* SLICE: As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
* MFA: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.

@@

```{r}
git2r::repository(".")
```

>>>>

```{r setup, include=FALSE}
source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/setup.R"))
```
`r glue::collapse(sample(c("Wouter Saelens* ¹ ²", "Robrecht Cannoodt* ¹ ² ³")), ", ")`, Helena Todorov¹ ² ⁴ ⁵, Yvan Saeys¹ ²

@@

Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes. Since 2014, more than 50 trajectory inference methods have been developed, each with its own set of methodological characteristics. This makes a direct[a] comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking. We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard.[b] Using this framework, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets. We employ several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness. We conclude with practical guidelines for method[c] users, and also believe our evaluation framework can be used to spearhead the development for new methods. Our evaluation pipeline is fully reproducible, can be easily extended, and is available at https://www.github.com/dynverse.

@@

Single-cell -omics technologies now make it possible to more accurately[d] model biological systems than ever before. One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation. When cells are sampled from a population in which cells are at different unknown points in the dynamic process,[e] trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016]. Because they offer an unbiased and transcriptome-wide understanding of the dynamic process[f], trajectories then allow the identification of new [g](primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs. This both for specific subsets of cells, but in the near future also for whole tissues or organisms[h] [@regev_scienceforumhuman_2017].

@@

Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`). Initially, most trajectory inference methods fixed the trajectory topology to a linear `r cite_methods(fixed_linear)`, bifurcating `r cite_methods(fixed_bifurcating)` or other `r cite_methods(fixed_other)` topology, by either algorithmic limitations `r cite_methods(fixed_algorithm)` or through user parameters `r cite_methods(fixed_parameters)`. The focus of these methods was therefore mainly to correctly ordering[i] the cells along this[j] fixed topology. However, several recent methods now infer both the ordering and the topology at the same time, although some topological limitations are still imposed (`r ref("sfig", "methods_over_time", "b")`). Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to[k] model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].

<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
Although trajectory inference methods use a variety of algorithms and steps to reach a final ordering, most consist of the following three main steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others. Furthermore, with the ongoing scalability in single-cell -omics technologies (reaching a million cells [@angerer_singlecellsmake_2017], it becomes more and more important that analysis methods become both scalable [@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user friendly.

<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem[l]. Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art. In this study, we therefore for the first time developed a[m] comprehensive evaluation framework for trajectory inference methods. We test[n] both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge. We include a rigorous parameter optimisation, making sure that parameters are optimized while avoiding overfitting on characteristics of specific datasets. To make our framework reusable and extendable, we make use of continuous analysis [@beaulieu-jones_reproducibility_2017], which allows the developers of new methods to easily test their methods and compare them against the state-of-the-art.

@@

We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "methods")`). We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characterisation")`).

@@

At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characterisation", "a")`). However, only a handful, mostly recent[o], methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).

@@

`r anchor("fig", "method_characterisation")`
![](`r figure_file("method_characteristics.svg", experiment_id="4-method_characterisation")`)
`r ref("fig", "method_characterisation", anchor=TRUE)` First characterization of each method.

@@

Most[p] methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characterisation", "b")`). Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`). However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characterisation", "e")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characterisation", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].

@@

The underlying algorithms can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building. Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "method_components")`). MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "method_components", "b")`).

@@

We compared the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "b")`). To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "e")` and `r ref("sfig", "wrapping")`. In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "c")`). This common trajectory format allowed us to compare the output of each method against gold standard trajectories. We compared the results of each TI method with both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of both datasets as best as possible (`r ref("fig", "evaluation_overview", "d")`). We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")`). We further divided the datasets into gold and silver standard, depending on whether the gold standard was extracted from only external knowledge (gold) or including the expression data itself (silver). To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "f")`). To calculate a final ordering, we combined metrics and datasets so that each metric, trajectory type and dataset source had equal weight.

```{r}
add_sfig({
                glue::glue("![]({figure_file('wrapping.svg', experiment_id = '4-method_characterisation')})")
}, 
"wrapping", 
"Our wrapping workflow", 
15, 
7.5
)
```

# Discussion
<!-- Some intro -->
Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data. Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list created by Sean Davis [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) . It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice[q]. In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.

@@

* DPT: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point. This was then processed as described above.
* Sincell: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.  This was then processed as described above.
* SLICER: Although this method returns a branch network and a global pseudotime, it was not clear to us how to assign cells to individual branches in the network. We therefore used the cell graph provided by the method to calculate the shortest paths between the start and end cells found by the method, and then used this graph as the backbone for all other cells, creating a local pseudotime for every cell within its branch. The output was then further processed as above.
* SLICE: As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
* MFA: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.

@@

```{r}
git2r::repository(".")
```
[a]This makes it difficult to directly compare between trajectory inference methods , and as a result (...)
[b]which facilitates the comparison of the results of trajectory inference methods (start using the TI abbreviation at some point?) between each other, and possibly to a gold standard.
[c]which might come in handy to users. We believe (...)
[d]move? "make it possible to model biological systems more accurately than ever before".
[e]When sampled, single cells will most probably be in different states of dynamic processes, due to the stochasticity of expression. Trajectory inference methods can be used to computationally order these cells along trajectories, based on gradual changes in the expression of several genes.
[f]processes?
[g]the trajectories can allow to identify new (primed) subsets of cells, to delineate the exact wiring of a differentiation tree and to infer the regulatory interactions which are responsible for a bifurcation.
[h]This is currently applicable to specific subsets of cells, but might also be applicable to whole tissues or organisms in the near future.
[i]order
[j]a?
[k]will be able to model any underlying structure in the data, without the need of user provided parameters. These methods might allow to model more complex behaviors, such as (...)
[l]get an idea of which available method or set of methods would optimally resolve their problem.
[m]we present the first comprehensive (...)?
[n]compared 22 TI methods on both synthetic (..)
[o]a handful of the most recent methods
[p]Repetition of most: put "The majority of" instead?
[q]analysis?
